--- 
title: " "
author: ""
date: ''
output: pdf_document
description: Introduction
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
---


# Access Economic Data via the BEA API

While BEA data can be accessed via the official website, programming-inclined users can easily access data through BEA's Application Programming Interface (API), which is a common way of querying and retrieving data. For R Users, BEA has developed the `bea.R` library that is optimized for econometricians, statisticians, and data analysts. `bea.R` can be installed directly from CRAN:

```{r, eval = FALSE}
install.packages("bea.R")
```

Upon installation, load the library:

```{r, message = FALSE, warning=FALSE}
library(bea.R)
```

In order for BEA to understand user needs and to ensure a consistent, high level of service, the BEA API requires an API key -- a simple 36 character code that opens access to data. The API key is free, asking for your name and email address: [https://www.bea.gov/API/signup/index.cfm](https://www.bea.gov/API/signup/index.cfm). Upon submitting an API key request, the key will be emailed to you.

Once you have received your BEA API key, save it to a variable to make it easier to use later:

```{r, eval = FALSE}
beaKey <- "[36-Character-Key]"
```
```{r, echo = FALSE}
#beaKey =  readLines("/Users/jeff/Documents/keys/beaKey.txt")
beaKey =  readLines("C:/Users/Jeff/Documents/beakey.txt")
```



## Searching for data 

Much like using an index in a textbook, making a successful data retrieval request to the API requires specific information (e.g. variable series code, frequency, geographic units). The R library's search functionality provides facility to easily surface the necessary information. Search is facilitated by the `beaSearch()` method, which requires two arguments: _keyword_ and _beaKey_. 

For example, the following keyword search returns a list of all datasets in which the term "Personal Consumption Expenditures" appears. One particularly important aspect of the search result is a pre-populated _API call_ that jump starts the querying process.

```{r, message = FALSE, warning=FALSE, eval = FALSE}
beaSearch('Personal Consumption Expenditures', beaKey)
```
```{r, message = FALSE, warning=FALSE, echo = FALSE}
knitr::kable(t(beaSearch('Personal Consumption Expenditures', beaKey)[1:2]), booktab = TRUE, caption = "Example results from through search interface.")
```

For some keyword searches, a sufficiently specific search will yield only a few results. Given the vast amount of data that BEA publishes, the search functionality will often return a sizable amount of candidate results. This may be more easily evaluated by specifying `asHtml = TRUE`, which presents search results in an interactive, searchable table.

```{r, message = FALSE, warning=FALSE, eval = FALSE}
beaSearch('Final sales of computers to domestic purchasers', beaKey, asHtml = TRUE)
```


## Get Data

Using the sample API call from the above example, we will retrieve _Personal Consumption Expenditures_ from _Table 1.1.1. Percent Change From Preceding Period in Real Gross Domestic Product_. But in short, it is `TableID = 1` from the `NIPA` dataset. Below, the list object specifies the relevant parameters, requesting the data at a quarterly frequency for all years.

```{r, message=FALSE, warning=FALSE}
specs <- list(
	'UserID' = beaKey,
	'Method' = 'GetData',
	'datasetname' = 'NIPA',
	'TableID' = '1',
	'Frequency' = 'Q',
	'Year' = '2010,2012,2013,2014',
	'ResultFormat' = 'json'
)

```

To retrieve a limited selection of multiple years, list all the years you want to retrieve. For example, to retrieve data for 2010 through 2014, use `"Year"="2010,2012,2013,2014"`.

The `beaGet()` function can use the `spec` list to retrieve data. By default, the data are returned in wide form -- each row is an indicator and each time period's data is represented as a separate column. 

```{r, message=FALSE, warning=FALSE}
df <- beaGet(specs);
head(df, 3)
```

For some use cases, in particular for time series models, obtaining the data in long form is desirable, which can be done by setting `asWide = FALSE`.

```{r, message=FALSE, warning=FALSE, eval = FALSE}
df <- beaGet(specs, asWide = FALSE);
head(df, 3)
```

In other cases, it may be more desirable to obtain the data in a form in which each column represents a variable, which is accomplished by setting `iTableStyle = FALSE`.

```{r, message=FALSE, warning=FALSE, eval = FALSE}
df <- beaGet(specs, iTableStyle = FALSE)
head(df, 3)
```

