--- 
title: "Maps"
date: '`r Sys.Date()`'
output: pdf_document
documentclass: book
site: bookdown::bookdown_site
biblio-style: apalike
---



# Interactive Maps

Interactive maps have become commonplace in the web's tools for communicating geographic information. While they are user friendly representations, building even the simplest of maps requires some basic knowledge of how the data are structured. For example, the map below plots the per capita personal income growth for 2016. 

```{r, message = FALSE, warning = FALSE, echo = FALSE, fig.cap = "What will be built in this session"}
library(bea.R)
library(leaflet)
library(sp)


#Personal Income Per Capita
pcinc_regional <- beaGet(
  list(
    userid   = beaKey, 
    method   = 'getdata', 
    linecode   = '3', 
    tablename  = 'CA1', 
    frequency  = 'a', 
    datasetname = 'regionalincome', 
    year     = 'all', 
    geofips   = 'state'
  )
)


pcinc_regional$pcinc.growth <- 100 * (pcinc_regional$DataValue_2016/pcinc_regional$DataValue_2015 - 1)
pcinc_regional$STATEFP <- substr(pcinc_regional$GeoFips,1,2)
pcinc_regional <- pcinc_regional[, c("STATEFP", "pcinc.growth")]


library(rgdal)

getShape <- function(url){
  # Download, unzip and load shapefile from URL
  #
  # Args: 
  #   url = URL to zip file containing shapefile
  #
  # Returns:
  #   
  #   Loaded shapefile
  
  #Create temp objects
  temp_dir <- tempdir()
  temp_file <- tempfile()
  
  #Download and unzip file
  download.file(url, temp_file,method="libcurl")
  unzip(temp_file, exdir = temp_dir)
  
  #Load shapefile
  file_prefix <- grep("*\\.shp$",list.files(temp_dir), value = TRUE)
  file_prefix <- gsub("\\.shp", "", file_prefix)
  return(readOGR(temp_dir, layer = file_prefix, verbose = FALSE))
}


# Get Shapefile
states <- getShape("http://www2.census.gov/geo/tiger/GENZ2016/shp/cb_2016_us_state_20m.zip")
states@data$STATEFP <- as.character(states@data$STATEFP)
states <- merge(states, pcinc_regional, by = "STATEFP")
states <- states[!is.na(states@data$pcinc.growth),]

## Setting Styles

library(RColorBrewer)

#Set up color palette
bins1 <- seq(floor(min(states@data$pcinc.growth)), ceiling(max(states@data$pcinc.growth)), 1)
pal1 <- colorBin("YlGnBu", domain = states@data$pcinc.growth, bins = bins1)



## Set labels

#Pop Up Label
labs <- sprintf(
  "<strong>%s</strong><br> <small>Per Capita Growth (2015-16)<br>PI:<span style='font-size:12px;color:darkblue;font-weight:bold'>%g </span> percent",
  states@data$NAME, round(states@data$pcinc.growth,2)) %>% lapply(htmltools::HTML)

#Set canvas
map_object <- leaflet(states, width = "100%", height = "500px") %>% 
  setView(-96.5, 37.7, 5) 


#Add background
map_object <- map_object  %>% 
  addProviderTiles(provider = "CartoDB.Positron") 


#Add data 
map_object <- map_object %>% 
  addPolygons(group = "Personal Income",
              fillColor = ~pal1(pcinc.growth), weight = 1, color = "white", fillOpacity = 0.5,
              highlight = highlightOptions(
                weight = 2, color = "#666", fillOpacity = 0.9),
              label = labs,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px", direction = "auto"))  
 

#Render
map_object

```

In order to render the color-coded polygons as a _chloropleth map_, we need two types of information: the _geometries_ of US states and their economic _attributes_. The shape of each state is stored in a special data format, such as a shapefile or GeoJSON, that contains  geographic coordinates and properties that indicate the type of shape (e.g. line, polygon, point). The geometries on their own are useful to render the outline of each state, but do not usually come furnished with economic data. Each shape is associated with an identifier that can be used to join new data, such as economic attributes.

Below, in order to construct a map containing growth of per capita personal income and personal expenditures by state, we layout and illustrate a five step process: 

1. Get BEA data from the API and make adjustments
2. Get US state shapefile from the US Census Bureau
3. Join the data
4. Set visual styles and functionality
5. Render map


## Get BEA data

Using the R library's `beaGet()` function, we will make two requests via the BEA API: 

- Annual Per Capita Personal Income by State as available in table _CA1_
- Annual Per Capita Personal Expenditure by State as available in table _C3_

In order to calculate growth, we will request two years of data: 2015 and 2016. In the code below, notice that the request parameters are first inserted in a `list()` object, then passed to the `beaGet()` function. The result is a data.table containing time series data.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(bea.R)

#beaKey =  readLines("/Users/jeff/Documents/keys/beaKey.txt")

#Personal Income Per Capita
  #Specify request parameters
  pcinc_req <- list(UserID   = beaKey, 
                    Method   = 'getdata', 
                    LineCode   = '1', 
                    TableName  = 'CA1', 
                    Frequencu  = 'a', 
                    DatasetName = 'regionalincome', 
                    Year     = '2015,2016', 
                    GeoFIPS   = 'state')
  
  #Send parameters via beaGet
  pcinc_regional <- beaGet(pcinc_req)
  
# Personal Consumption (IndustryID = 1 indicates all)
  #Specify request parameters
  pcpce_req <- list(UserID = beaKey, 
                    Method = 'GetData', 
                    DatasetName = 'RegionalProduct', 
                    Component = 'PCPCE_SAN',
                    IndustryID = '1' , 
                    Year = '2015,2016', 
                    GeoFIPS = 'state', 
                    Frequency = 'a')
  
  #Send parameters via beaGet
  pcpce <- beaGet(pcpce_req)

```
```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(bea.R)

beaKey =  "36-character key goes here"

#Personal Income Per Capita
  #Specify request parameters
  pcinc_req <- list(UserID   = beaKey, 
                    Method   = 'getdata', 
                    LineCode   = '1', 
                    TableName  = 'CA1', 
                    Frequencu  = 'a', 
                    DatasetName = 'regionalincome', 
                    Year     = '2015,2016', 
                    GeoFIPS   = 'state')
  
  #Send parameters via beaGet
  pcinc_regional <- beaGet(pcinc_req)
  
# Personal Consumption (IndustryID = 1 indicates all)
  #Specify request parameters
  pcpce_req <- list(UserID = beaKey, 
                    Method = 'GetData', 
                    DatasetName = 'RegionalProduct', 
                    Component = 'PCPCE_SAN',
                    IndustryID = '1' , 
                    Year = '2015,2016', 
                    GeoFIPS = 'state', 
                    Frequency = 'a')
  
  #Send parameters via beaGet
  pcpce <- beaGet(pcpce_req)
```



## Get shapefile

Next, a geometry file is required. The US Census Bureau publishes shapefiles for select geographic areas, including US states. Boundary files can be downloaded from [https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html). 

BEA has developed a simple function to download, unzip, and load Census shapefiles. This function requires the `rgdal` package to load shapefiles.

```{r, message = FALSE, warning = FALSE}

library(rgdal)
getShape <- function(url){
  #
  # Download, unzip and load shapefile from URL
  #
  # Args: 
  #   url = URL to zip file containing shapefile
  #
  # Returns:
  #   
  #   Loaded shapefile
  
  #Create temp objects
  temp_dir <- tempdir()
  temp_file <- tempfile()
  
  #Download and unzip file
  download.file(url, temp_file,method="libcurl")
  unzip(temp_file, exdir = temp_dir)
  
  #Load shapefile
  file_prefix <- grep("*\\.shp$",list.files(temp_dir), value = TRUE)
  file_prefix <- gsub("\\.shp", "", file_prefix)
  return(readOGR(temp_dir, layer = file_prefix, verbose = FALSE))
}
```

Upon loading the function, we now can import a US state shapefile direct from the Census website.

```{r, message=FALSE, warning=FALSE}
url <- "http://www2.census.gov/geo/tiger/GENZ2016/shp/cb_2016_us_state_20m.zip"
states <- getShape(url)
```  

In its most basic form, the shapefile yields a map of state outlines. 

```{r}
plot(states)
```

## Process and join data

With the shapefile and indicators loaded, we can inspect the layout using the `head()` function. From the geometry side (e.g. `states`), the `STATEFP` field is a Federal Information Processing (FIPS) code, a hierarchical classification system to uniquely identify geographic units. This field will be used to match between the shapefile and each of the indicator files. 
```{r, echo = FALSE}
head(states@data, 5)
```

On the indicator side (e.g. `pcpce` and `pc_regional`), the `GeoFips` field will be required to merge with the geometry file and the the `DataValue_` fields will be converted into annual percentage growths. The `GeoFips` field will need to be changed to match the `STATEFP` format by extracting only the first two digits of the identifier.

```{r, eval = FALSE}
head(pcpce, 5)
```
```{r, echo = FALSE, fig.cap = "Sample extract from the Per Capita PCE data"}
knitr::kable(head(pcpce,5), booktab = TRUE)

```

__Data Manipulation__
To calculate the annual growth, we simply use the formula:

$$\Delta y_t = 100 \times (\frac{y_t}{y_{t-1}}-1) = 100 \times (\frac{\text{DataValue_2016}}{\text{DataValue_2015}}-1) $$

In addition, the `GeoFips` code can be shortened to the first two digits using the `substr()` function. The `attach()` function is used so that the name of variables within a specified dataframe are treated as global variables, removing the need to specify the data frame each time a variable is called and allowing the code to be more compact. 

```{r}
#Calculate Per Capita PCE Growth
  attach(pcpce)
  pcpce$pcpce.growth <- 100 * (DataValue_2016/DataValue_2015 - 1)
  pcpce$STATEFP <- substr(GeoFips,1,2)
  pcpce <- pcpce[, c("STATEFP", "pcpce.growth")]
  detach(pcpce)

#Calculate Per Capita PCE Growth
  attach(pcinc_regional)
  pcinc_regional$pcinc.growth <- 100 * (DataValue_2016 / DataValue_2015 - 1)
  pcinc_regional$STATEFP <- substr(GeoFips,1,2)
  pcinc_regional <- pcinc_regional[, c("STATEFP", "pcinc.growth")]
  detach(pcinc_regional)
```


__Merge__
With the data processed, we will use the `merge()` function to join the economic data frames together using the newly created `STATEFP` variable. Then this merged data frame is combined with the `states` shapefile. 
```{r}
#Merge both per capita data sets
  pcinc_regional <- merge(pcinc_regional, pcpce, by = "STATEFP")
  
#Merge shapefile to the data set
  states <- merge(states, pcinc_regional, by = "STATEFP")
  states <- states[!is.na(states@data$pcinc.growth),]
```

## Set styles
A color palette needs to be defined in order for colors to be rendered in a map. Color palettes require two pieces of information: number of bins and color progression.  

- Bins are defined by the cutoffs in a specified variable. Below, we use the `seq()` function to produce a sequence of equally spaced thresholds between a minimum value (`min()`) and maximum value (`max()`).  `floor()` and `ceiling()` are used to round values to the next smallest and next largest number, respectively.
- The `colorBin()` function available in the `leaflet` library creates color palettes using three parameters: a color palette (e.g. "Reds", "Greens"), a domain (e.g. a variable like `pcinc.growth`), and the thresholds that define the bins (defined above). Each `pcinc.growth` and `pcpce.growth` will have their own palletes labeled `pal1` and `pal2`, respectively.

```{r, message = FALSE, warning = FALSE}
#Load leaflet library
  library(leaflet)

#Set upbins
  attach(states@data)
  bins1 <- seq(floor(min(pcinc.growth)), 
               ceiling(max(pcinc.growth)), 1)
  pal1 <- colorBin("Blues", domain = pcinc.growth, bins = bins1)
  
  bins2 <- seq(floor(min(pcpce.growth)), 
               ceiling(max(pcpce.growth)), 1)
  pal2 <- colorBin("Reds", domain = pcpce.growth, bins = bins2)
  detach(states@data)
```

Most maps also tend to be furnished with tooltips, which are a labels or comments that popup when a mouse/cursor moves over objects in the map. The `sprintf()` function helps to blend text and numbers as contained in variables. Below, simple HTML text is used to create three lines in which text and numeric values are populated where `%s` and `%g` appear, respectively.

```{r, message = FALSE, warning = FALSE}
#Pop Up Label
labs <- sprintf(
  "<strong>%s</strong><br> 
  PI: %g percent <br>
  PCE: %g percent",
  states@data$NAME, round(states@data$pcinc.growth,2), round(states@data$pcpce.growth,2)) %>% 
  lapply(htmltools::HTML)
```

## Render 
With all the settings sorted, we now can begin to put together the map itself. The map will be furnished with four attributes: 

- A map canvas that is centered on the middle of the US
- A base layer that provides context
- Two layers of state polygons that are color coded based on each `pcinc.growth` and `pcpce.growth`
- Toggles for the polygons

To start, we will create a map object that specifies the shapefile and the canvas extent. To that object we use the `%>%` operator to set the view by centering the map on longitude = -96.5 and latitude = 37.7. Notice that the rendered object is blank as we have not yet specified the base layer or polygons.
```{r, message = FALSE, warning = FALSE}
#Set canvas
map_object <- leaflet(states, width = "100%", height = "500px") %>% 
              setView(-96.5, 37.7, 4) 
```
```{r, echo = FALSE, fig.height=1.5}
map_object
```

Next, we add web map tiles from CartoDB using the `addProviderTiles`. Browse other free tiles at [http://leaflet-extras.github.io/leaflet-providers/preview/](http://leaflet-extras.github.io/leaflet-providers/preview/).
```{r, message=FALSE, warning=FALSE}
#Add background
map_object <- map_object  %>% 
              addProviderTiles(provider = "CartoDB.Positron") 
```
```{r, echo = FALSE}
map_object
```


Next, we at the polygons, which is a fairly verbose step. But in short: 

- For each polygon, we associate it with a `group` that will later be used in the toggle controls. 
- The `fillColor` is derived from the previously specified palette objects. 
- The `weight` and `color` indicate the thickness and border color of each polygon.
- The `fillOpacity` indicates how transparent the polygon shading will be (between 0 and 1).
- `highlightOptions` indicate how each polygon's colors will change when a cursor moves over.
- `label` uses the `labs` object to create the tooltip popup.
- `labelOptions` control how the tooltip looks.

```{r}
#Add data 
map_object <- map_object %>% 
  
                addPolygons(group = "Personal Income",
                  fillColor = ~pal1(pcinc.growth), weight = 1, 
                  color = "white", fillOpacity = 0.5,
                  highlight = highlightOptions(
                    weight = 2, color = "#666", fillOpacity = 0.4),
                  label = labs,
                  labelOptions = labelOptions(
                    style = list("font-weight" = "normal", padding = "3px 8px"),
                    textsize = "15px", direction = "auto")) %>% 
  
                addPolygons(group = "Personal Consumption",
                  fillColor = ~pal2(pcpce.growth), weight = 1, 
                  color = "white", fillOpacity = 0.5,
                  highlight = highlightOptions(
                    weight = 2, color = "#666", fillOpacity = 0.4),
                  label = labs,
                  labelOptions = labelOptions(
                    style = list("font-weight" = "normal", padding = "3px 8px"),
                    textsize = "13px")) 
```
```{r, echo = FALSE}
map_object
```

Lastly, `addLayersControl` are added to the map object to define the polygon layers that can be toggled.
```{r, message = FALSE, warning = FALSE}
map_object <- map_object %>%
              addLayersControl(
                overlayGroups = c("Personal Income", "Personal Consumption"))
```
```{r, echo = FALSE}
map_object
```


The result is a fully functional interactive map that can help build and support narratives on the economy over space.
